Galaxy Zoo: the dependence of the star formation-stellar mass relation on spiral disk morphology

### 2015-02-04 paper/mainsequence_morph.tex
Completed draft 2 with new mass limits

Stellar masses and star formation rates are computed from optical diagnostics and taken from the MPA-JHU catalogue \citep{kau03a,bri04,sal07}. We use updated masses and activity classifications from the DR7 database.\footnote{\url{http://home.strw.leidenuniv.nl/\mytilde jarle/SDSS/}} We select only galaxies with $M_\star > 10^8 M_\odot$, for which \citet{bri04} estimate that the MPA-JHU sample is complete. Star-forming galaxies are separated from other emission-line galaxies using the standard BPT classification \citep*{bal81} below the \citet{kau03} demarcation. Galaxies classified as star-forming but with low signal-to-noise ratio $(S/N > 3)$ are also excluded. Both $M_\star$ and $SFR$ are measured using median values extracted from the PDF. % Footnote is breaking LaTeX.

@@

Two-armed spiral galaxies and those with uncertain multiplicities are the only morphologies that extend slightly below the SFMS. One possibility is that this is a consequence of galaxies that have already begun the process of quenching. As a result, the star-formation rate is depleted and the contrast of spiral arms against the rest of the disk (which for visual identification is at least in part dependent on bright knots of recent star formation) makes the multiplicity more difficult to identify. However, the $(u-r)$ colors of these galaxies (uncorrected for dust extinction) are on average significantly bluer than green-valley late-types as identified in \citet{sch14}.

We have repeated the analysis above for the sub-sample of disk galaxies for which the spiral multiplicity is determined with high confidence ($p_\textrm{arms~number} > 0.8$) by GZ2, thus eliminating ``intermediate'' galaxies for which the morphology is uncertain. These galaxies ($N=14,148$) are dominated by two-armed spirals, which are the only spiral multiplicity for which significant numbers of galaxies at $M_\star<10^9~M_\odot$ are detected. The results for the SFMS are qualitatively the same as when using the weighting scheme, although we note that there are too few examples ($N<10$) of either three- or four-armed spirals for a reliable fit. The offset of the one-armed spirals above the SFMS is also significantly more pronounced when using high-confidence morphologies.

@@

It should also be noted that the galaxies in GZ2 flagged as a function of pitch angle are not representative of the true vote distribution. The raw numbers in Figure~\ref{fig-winding} suggest that most spiral galaxies are either tightly or loosely wound. In fact, the plurality classification for most galaxies is for medium-winding; the spread in votes is typically large, though, and so users rarely agree on the medium option at the 80\% level which sets the flag. An alternative method is to analyze the morphology of spiral galaxies by directly weighting them as a function of the pitch angle categories (Figure~\ref{fig-winding_weighted}), which has the advantage of including all spiral galaxies. This shows an even tighter agreement between the samples separated by pitch angle and that of the full star-forming sample.

@@

The agreement of all sub-varieties of star-forming galaxies is supported by the close agreement to the linear fits to the data for all well-sampled categories (Table~\ref{tbl-fits}). This tracks only the slope and offset of the distribution, however, and not its width. We thus also compare the sample standard deviation ($\sigma_{SFR}$) to the star-forming galaxy population over its various morphological subsamples. The value of $\sigma_{SFR}$ monotonically decreases with increasing mass over the range $6.0<\log(M/M_\odot)<11.5$. For all populations examined in this paper, though, the widths of their distributions are consistent with the broader population (Figure~\ref{fig-sigma}).

@@

We have examined all the populations of galaxies described above (bars, arm pitch angle, arm multiplicity) and measured the differences when using specific star formation rate ($\textrm{sSFR}\equiv\textrm{SFR}/M_\star$) instead of SFR. There is no significant change in any of the results.

>>>>

Stellar masses and star formation rates are computed from optical diagnostics and taken from the MPA-JHU catalogue \citep{kau03a,bri04,sal07}. We use updated masses and activity classifications from the DR7 database.\footnote{\url{http://home.strw.leidenuniv.nl/\mytilde jarle/SDSS/}} We select only galaxies with $M_\star > 10^8 M_\odot$, for which \citet{bri04} estimate that the MPA-JHU sample is complete. Star-forming galaxies are separated from other emission-line galaxies using the standard BPT classification \citep*{bal81} below the \citet{kau03} demarcation. Galaxies classified as star-forming but with low signal-to-noise ratio $(S/N > 3)$ are also excluded. Both $M_\star$ and SFR are measured using median values extracted from the PDF. % Footnote is breaking LaTeX.

@@

The only morphologies that extend slightly below the SFMS are those with the highest level of multiplicity (five or more arms). The best-fit line for this population has a steeper slope, driven by the galaxies with relatively low SFR at $10^9<M/M_\star<10^{10}$. This is a new and unusual result; one possible explanation is that stellar disks in the process of quenching will have fewer bright \mbox{H\,{\sc ii}}~regions and the contrast between the arm and interarm regions is increased. This could result in better visibility for older (and potentially overlapping) spatial modes in the galaxy's disk, increasing the measured multiplicity. It should be emphasized, though, that 5+-arm spirals represent the smallest morphological group in the sample, and that the associated fit errors in Table~\ref{tbl-fits} are the largest for any multiplicity; simple statistical variance cannot be ruled out as an explanation for the best-fit line to the SFMS.

%One possibility is that this is a consequence of galaxies that have already begun the process of quenching. As a result, the star-formation rate is depleted and the contrast of spiral arms against the rest of the disk (which for visual identification is at least in part dependent on bright knots of recent star formation) makes the multiplicity more difficult to identify. However, the $(u-r)$ colors of these galaxies (uncorrected for dust extinction) are on average significantly bluer than green-valley late-types as identified in \citet{sch14}.

We have repeated the analysis above for the sub-sample of disk galaxies for which the spiral multiplicity is determined with high confidence ($p_\textrm{arms~number} > 0.8$) by GZ2, thus eliminating ``intermediate'' galaxies for which the morphology is uncertain. These galaxies ($N=10,035$) are dominated by two-armed spirals, which are the only spiral multiplicity for which significant numbers of galaxies at $M_\star<10^9~M_\odot$ are detected. The results for the SFMS are qualitatively the same as when using the weighting scheme, although we note that there are too few examples ($N<10$) of either three- or four-armed spirals for a reliable fit. The offset of the one-armed spirals above the SFMS is also significantly more pronounced when using high-confidence morphologies.

@@

It should also be noted that the galaxies in GZ2 flagged as a function of pitch angle are not representative of the true vote distribution. The points in Figure~\ref{fig-winding} would suggest that there are relatively few spiral galaxies overall, and that most are either tightly or loosely wound. In fact, the plurality classification for most galaxies is for medium-winding; the spread in votes is typically large, though, and so users rarely agree on the ``medium'' option at the 80\% level which sets the flag. An alternative method is to analyze the morphology of spiral galaxies by directly weighting them as a function of the pitch angle categories (Figure~\ref{fig-winding_weighted}), which has the advantage of including all spiral galaxies. This shows an even tighter agreement between the samples separated by pitch angle and that of the full star-forming sample.

@@

The agreement of all sub-varieties of star-forming galaxies is supported by the close agreement to the linear fits to the data for all well-sampled categories (Table~\ref{tbl-fits}). This tracks only the slope and offset of the distribution, however, and not its width. We thus also compare the sample standard deviation ($\sigma_{SFR}$) to the star-forming galaxy population over its various morphological subsamples. The value of $\sigma_{SFR}$ monotonically decreases with increasing mass over the range $8.0<\log(M/M_\odot)<11.5$. For all morphological populations examined in this paper, the widths of their distributions are consistent with the broader population (Figure~\ref{fig-sigma}).

@@

We have also examined all the populations of galaxies described above (bars, arm pitch angle, arm multiplicity) and measured the differences when using specific star formation rate ($\textrm{sSFR}\equiv\textrm{SFR}/M_\star$) instead of SFR. There is no significant change in any of the results.
